Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-21 Dail_Mad_ext.tex
first commit

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Accounting for
process variation and observation error in such efforts is necessary
to achieve unbiasedness, and doing so is possible using recently
developed state-space models.
%for estimators of population change to be unbiased, and state-space
%models have been developed for this
%models are widely used for these purposes because they allow for
%estimating parameters of classical growth models, such as exponential,
%logistic, and Gompertz, while accounting for demographic stochasticity
%and observation error.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
We expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the  \textbf{R} package
\texttt{unmarked} or
\textbf{JAGS} is also provided.
%is provided for doing so.
%Furthermore, we describe classical and
%Bayesian methods of parameter estimation, and we present methods of
%forecasting population size under future environmental
%conditions.
%These new developments will
%allow researchers to apply these methods to address important
%questions regarding the factors affecting spatial and temporal
%variation in abundance.
A simulation study indicated that bias was negligible and coverage nominal
and an analysis of North American Breeding Bird Survey is presented
for illustration.
}

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In this paper, we focus on the model of
\citet{dail_madsen:2011} %developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficieinces with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frquentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variaion of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

>>>>

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Accounting for
process variation and observation error in such efforts is necessary
to achieve unbiasedness \hl{[may be too strange sounding a term for
  abstract]}, and doing so is possible using recently
developed state-space models.
%for estimators of population change to be unbiased, and state-space
%models have been developed for this
%models are widely used for these purposes because they allow for
%estimating parameters of classical growth models, such as exponential,
%logistic, and Gompertz, while accounting for demographic stochasticity
%and observation error.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
We expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the  \textbf{R} package
\texttt{unmarked} or
\textbf{JAGS} is also provided.
%is provided for doing so.
%Furthermore, we describe classical and
%Bayesian methods of parameter estimation, and we present methods of
%forecasting population size under future environmental
%conditions.
%These new developments will
%allow researchers to apply these methods to address important
%questions regarding the factors affecting spatial and temporal
%variation in abundance.
A simulation study indicated that bias was negligible and coverage nominal
and an analysis of North American Breeding Bird Survey is presented
for illustration.
}

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation: \hl{[I'm not familiar with bracket notation,
  so these equations are confusing to me.  How widely used is this
  notation in ecology?  Will others likely understand this?]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In this paper, we focus on the model of
\citet[henceforth the DM model;]{dail_madsen:2011}
%developed a model (henceforth the DM model)
that simultaneous resolves each of the abovementioned problems with
traditional state-space models, and is designed for simple count
data.
\begin{comment}
  The model (1) recognizes
  the discrete nature of count data and the underlying true abundance
  state, (2) includes a realistic observation model with explicit
  detection probability parameters (3) has estimable parameters,
  including those describing the initial state at time $t=1$, and (4)
  allows for inference about both temporal and spatial variation in
  abundance.

  In some respects, the DM model can be viewed as a specific
  manifestation of the general state-space model formulated by
  \citep{buckland_etal:2004}. However, their model was designed to
  allow for inference about demographic parameters, which will
  typically require mark-recapture data. Although this may be viewed
  as the gold standard in many studies of wildlife populations, in
  this paper, we focus on the analysis of count data, which is the
  typical source of information used in state-state modeling
  applications, and one of the most widely available forms of
  ecological data---it is also the data required by the DM model.
\end{comment}
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accomodate classical models of population growth and to
handle several features common to ecological
time-series. Specifically, we describe methods for accomodating excess
zeros and nuisance variables such as random observer effects. Both
frequentist and Bayesian methods of inference are discussed, and code
for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accomodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of  the paper is to provide ecologists with flexible and accessible means of
adressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
